---
title: "DS202-FinalProject"
author: "Corbin Graham, Coleman Dimmer"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DS 202 Final Project

**Title: Climate Analysis of Milan, Italy**

**Authors: Corbin Graham, Coleman Dimmer**

The goal of this project is to demonstrate skills learned, in R, analysis of data, demonstration of data, and drawing conclusions from data. We used data from National Centers for Environmental Information (NCEI) for the Milan Weather Recording Station of Milan, Italy. Using graphs, charts, and analysis, we will visualize and demonstrate patterns and trends within the dataset.

## Introduction

The United Nations is an intergovernmental organization whose stated purposes are to maintain international peace and security. They describe themselves as, \"the one place on Earth where all the world\'s nations can gather together, discuss common problems, and find shared solutions that benefit all of humanity.\"

The United Nations, as part of its mission for international peace, has compiled 17 goals, which it deems the most important and of the highest priority for all nations. The goals are ordered by importance to the organization\'s purpose. However, all goals should be considered of highest importance for every nation.

The United Nations\' thirteenth goal is on climate action. They state that: \"Every person, in every country in every continent will be impacted in some shape or form by climate change. There is a climate cataclysm looming, and we are underprepared for what this could mean.\" Thus, this can be seen as a topic of very high importance since, according to the United Nations, climate change will affect us all.

Finding accurate climate data over an extended period can be difficult since most countries did not begin practicing accurate data recording of weather patterns, emissions, until sometime within the last 200 years. For demonstrating, and finding accurate trends, it is important to look at accurate, comprehensive data collection. The most reliable source for climate data available in the United States is the United States National Centers for Environmental Information (NCEI). All the data they provide is extensive, and heavily scrutinized. The rigor of data quality provides us with the best, and most accurate measurements of the climate. Because most quality data recording began recently, we must look outside our borders to find older data with the same accuracy. The longest-running data record on climate comes from Milan, Italy, which began recording weather and climate data in 1763 for the Italian government.

Using this data, it is our goal to demonstrate climate patterns, and find trends in those patterns. Using the trends, we may, or may not find, we hope to find the highest correlating data available to potentially explain the trends we see. While some trends may have no correlation, it is of high likelihood that any trends demonstrated in climate patterns are the result of some external event.

## Data Background

The dataset we chose is monthly climate data recorded in Milan, Italy, between January 1763 and November 2008. We can use this data to analyze climate patterns. etc.

The original dataset has different information regarding both temperature and precipitation data, with further on more information about specific weather observations in the month.

**Degree Days**

-   Degree Days: days in a month where the temperature is above or below the mean temperature for that month, measured in points where each point is a degree above or below the mean.

-   Degree Cooling Days: days of the month with a temperature below that month's mean

-   Degree Heating Days: days of the month with a temperature above that month's mean

**Temperature**

-   Min/Max/Mean temperatures, with days the mins/maxes appeared on

-   Heating/Cooling Degree Days The amount of days above/below certain temperature levels

**Precipitation**

-   Total amount observed

-   Greatest amount observed with date Snow/Sleet total fall, max depth, max date Whether there was heavy fog/thunderstorms

#### Dataset

[Milan Weather Recording Station from NCEI (https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-month?startDate=1800-01-03T00:00:00&endDate=1800-01-03T23:59:59)](https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-month?startDate=1800-01-03T00:00:00&endDate=1800-01-03T23:59:59)

#### Data Columns

-   Extreme maximum precipitation
-   Extreme maximum temperature
-   Precipitation
-   Number days with maximum temperature greater than 90F (32.2C)
-   Number days with greater than 0.10 inch (2.54mm) of precipitation
-   Heating Degree Days Season to TIME
-   Heating Degree Days
-   Number days with maximum temperature greater than 70F (21.1C)
-   Cooling Degree Days Season to Date
-   Number days with greater than 0.01 inch (0.25mm) of precipitation
-   Day of month of highest daily total of precipitation
-   Extreme minimum temperature
-   Number days with minimum temperature less than 32F (0C)
-   Number days with minimum temperature less than 0F (-17.6C)
-   Day of month with extreme maximum temperature
-   Number days with maximum temperature less than 32F (0C)
-   Number days with greater than 1.00 inch (25.4mm) of precipitation
-   Cooling Degree Days
-   Average Maximum Temperature
-   Average Average Temperature
-   Average Minimum Temperature
-   Day of month of extreme minimum temperature

## Importing Data

Now that we have determined the dataset we wish to use, it is as simple as downloading it from the official source found to the CSV format. This allows us to easily import it as a malleable object.

#### Importanting Libraries

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

The libraries we are using are for inline operations, and plotting.

#### Important Data from CSV

```{r}
data <- read_csv("milan.csv")
head(data)
```

Since we have already downloaded our data to a CSV file, we only need to import it from that file.

#### Data Time Frame

```{r}
min(data$DATE)
max(data$DATE)
```

This function returns the first, and last date in the dataset. The first date is January, 1763, and the final date is November, 2008.

#### Data Columns

```{r}
colnames(data)
```

This function prints out all of the column names in the original dataset.

## Cleaning Data

Now that we have successfully imported the data, we will clean up our data object so our results are more accurate, and understandable.

#### Finding Important Variables

To 'weed out' invaluable variables, we will search for variables that lack the proper time frame, are null, or incomplete.

```{r}
# Determining if the location value changes
print("Latitude:")
unique(data$LATITUDE)
print("Longitude:")
unique(data$LONGITUDE)

# Determine if all variables have the same time frame
print("Number of NA values per column:")
colSums(is.na(data))

# As seen, DP values have quite a few more NA values than DT/DX values.
```

#### Creating a Clean Object

Drop null values, drop precipitation data (due to lack of data within timeframe), and remove location data since all of the recordings happened from the same location.

```{r}

#Grabs data and names columns to more reasonable col names. NOTE: DX and DY values are in degrees F, so changed to more accurate degrees C titles to match all other vars. 
milanClimate <- data %>%
  select(
    DATE,
    MEAN = TAVG,
    MEAN_MAX = TMAX,
    MEAN_MIN = TMIN,
    HEAT_DEG_DAYS = HTDD,
    COOL_DEG_DAYS = CLDD,
    HIGHEST_TEMP = EMXT,
    HIGHEST_DAY = EMXT_ATTRIBUTES,
    COOLEST_TEMP = EMNT,
    COOLEST_DAY = EMNT_ATTRIBUTES,
    GTE_32 = DX90,
    GTE_21 = DX70,
    GTE_00 = DX32,
    LTE_00 = DT32,
    LTE_N18 = DT00
  )

milanClimate$DATE <- lubridate::ym(milanClimate$DATE)

milanClimate$HIGHEST_DAY <- as.numeric(substr(gsub('\\D','',milanClimate$HIGHEST_DAY), nchar(gsub('\\D','',milanClimate$HIGHEST_DAY))-1, nchar(gsub('\\D','',milanClimate$HIGHEST_DAY))))

milanClimate$COOLEST_DAY <- as.numeric(substr(gsub('\\D','',milanClimate$COOLEST_DAY), nchar(gsub('\\D','',milanClimate$COOLEST_DAY))-1, nchar(gsub('\\D','',milanClimate$COOLEST_DAY))))

colnames(milanClimate)
```

## Marginal Summaries

Now that we have a cleaned data object, we can begin to plot charts to find and demonstrate different trends.

### Max Temperature and Minimum Temperature with Date

```{r}

subset_min <- milanClimate %>%
  filter(MEAN == min(MEAN, na.rm = TRUE)) %>%
  select(DATE, MEAN, MEAN_MIN, MEAN_MAX)
subset_min

subset_max <- milanClimate %>%
  filter(MEAN == max(MEAN, na.rm = TRUE)) %>%
  select(DATE, MEAN, MEAN_MIN, MEAN_MAX)
subset_max

subset_min_max <- bind_rows(subset_min, subset_max)
subset_min_max # Highest and Lowest Temperatures

ggplot(milanClimate, aes(x=DATE, y=MEAN)) +
  geom_bar(stat="identity", fill="blue", width=0.5) +
  geom_errorbar(aes(ymin = MEAN_MIN, ymax = MEAN_MAX), width = 0.2, position = position_dodge(width = 0.5)) +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Temperature Ranges by Date")

```

```{r}
# Monthly averages
ggplot(milanClimate, aes(x=DATE, y=MEAN)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Temperature by Date with Trend")
```

```{r}
# Extract year from the "date" column
new_data <- milanClimate %>%
  mutate(year = lubridate::year(DATE))

# Create a summary dataframe with yearly statistics
summary_df <- new_data %>%
  group_by(year) %>%
  summarise(
    MEAN = mean(MEAN, na.rm = TRUE),
    MIN = min(MEAN_MIN, na.rm = TRUE),
    MAX = max(MEAN_MAX, na.rm = TRUE)
  )
summary_df

# Print the summary dataframe
ggplot(summary_df, aes(x=year, y=MEAN)) +
  geom_bar(stat="identity", fill="blue", width=0.5) +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Annual Temperature Range with Trend")
```

```{r}

ggplot(milanClimate, aes(x=DATE, y=MEAN_MAX)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Maximum Temperature Annualy")

ggplot(milanClimate, aes(x=DATE, y=MEAN_MIN)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Minimum Temperature Annualy")

ggplot(milanClimate, aes(x=DATE, y=HEAT_DEG_DAYS)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Heating Days")

ggplot(milanClimate, aes(x=DATE, y=COOL_DEG_DAYS)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Cooling Days")

ggplot(milanClimate, aes(x=DATE, y=GTE_32)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Temperatures about 32C")

```

## 2 Plans for Data Exploration

With the data, we plan to explore the shift of temperatures in the Milan area. Combining this with economic data from the area can share how the shift in temperature has aftected the economy. Or combining it with populaiton data, we can compare it with population shifts. Across the data, we should be able to find comparisons between the temperature and other local events.
