---
title: "DS202-FinalProject"
author: "Corbin Graham, Coleman Dimmer"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DS 201 Final Project

**Authors: Corbin Graham, Coleman Dimmer**

## Description

The goal of this project is to demonstrate climate trends related to temperature throughout time. We are using data provided by National Centers for Environmental Information (NCEI) for the Milan weather recording station of Milan, Italy. Using graphs, charts, and statistical analysis, we will be able to visualize and demonstrate climate trends for this location.

We chose this location due to the timeframe and data provided. We will be using monthly weather (temperature) data from January 1763 to November 2008.

## Introduction

\-- Introduction Here --

\^Clearly and concisely describes the topic, and why it is of interest. Sets up a clear roadmap for the rest of the paper.

### Data Background

\-- Information about the data --

The dataset we chose is monthly climate data recorded in Milan, Italy, between January 1763 and November 2008. We can use this data to analyze climate patterns. etc.

The original data set has different information regarding both temperature and precipitation data, with further on more information about specific weather observations in the month.

Before going more in depth into what the data describes, let's talk some terminology:
  Degree days - the difference between the daily temperature mean and monthly temperature mean
    Cooling: More days below monthly average
    Heating: More days above monthly average


For temperature-based measurements, we have:
  Min/Max/Mean temperatures, with days the mins/maxes appeared on
  Heating/Cooling Degree Days
  The amount of days above/below certain temperature levels
  
For precipitation-based and weather-based data, we have:
  Total amount
  greatest amount observed with date
  Snow/Sleet total fall, max depth, max date
  Whether there was heavy fog/thunderstorms
  
  

\^Clearly and concisely describes the data and all relevant variables

#### Dataset

[Milan Weather Recording Station from NCEI (https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-month?startDate=1800-01-03T00:00:00&endDate=1800-01-03T23:59:59)](https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-month?startDate=1800-01-03T00:00:00&endDate=1800-01-03T23:59:59)

#### Data Columns

-   Extreme maximum precipitation
-   Extreme maximum temperature
-   Precipitation
-   Number days with maximum temperature greater than 90F (32.2C)
-   Number days with greater than 0.10 inch (2.54mm) of precipitation
-   Heating Degree Days Season to TIME
-   Heating Degree Days
-   Number days with maximum temperature greater than 70F (21.1C)
-   Cooling Degree Days Season to Date
-   Number days with greater than 0.01 inch (0.25mm) of precipitation
-   Day of month of highest daily total of precipitation
-   Extreme minimum temperature
-   Number days with minimum temperature less than 32F (0C)
-   Number days with minimum temperature less than 0F (-17.6C)
-   Day of month with extreme maximum temperature
-   Number days with maximum temperature less than 32F (0C)
-   Number days with greater than 1.00 inch (25.4mm) of precipitation
-   Cooling Degree Days
-   Average Maximum Temperature
-   Average Average Temperature
-   Average Minimum Temperature
-   Day of month of extreme minimum temperature

## Importing Data

Now that we have determined the dataset we wish to use, it is as simple as downloading it from the official source found to the CSV format. This allows us to easily import it as a malleable object.

#### Importanting Libraries

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

The libraries we are using are for inline operations, and plotting.

#### Important Data from CSV

```{r}
data <- read_csv("milan.csv")
head(data)
```

Since we have already downloaded our data to a CSV file, we only need to import it from that file.

#### Data Time Frame

```{r}
min(data$DATE)
max(data$DATE)
```

This function returns the first, and last date in the dataset. The first date is January, 1763, and the final date is November, 2008.

#### Data Columns

```{r}
colnames(data)
```

This function prints out all of the column names in the original dataset.

## Data Cleaning

Now that we have successfully imported the data, we will clean up our data object so our results are more accurate, and understandable.

#### Finding Important Variables

To 'weed out' invaluable variables, we will search for variables that lack the proper time frame, are null, or incomplete.

```{r}
# Determining if the location value changes
print("Latitude:")
unique(data$LATITUDE)
print("Longitude:")
unique(data$LONGITUDE)

# Determine if all variables have the same time frame
print("Number of NA values per column:")
colSums(is.na(data))

# As seen, DP values have quite a few more NA values than DT/DX values.
```

#### Creating a Clean Object

Drop null values, drop precipitation data (due to lack of data within timeframe), and remove location data since all of the recordings happened from the same location.

```{r}

#Grabs data and names columns to more reasonable col names. NOTE: DX and DY values are in degrees F, so changed to more accurate degrees C titles to match all other vars. 
milanClimate <- data %>%
  select(
    DATE,
    MEAN = TAVG,
    MEAN_MAX = TMAX,
    MEAN_MIN = TMIN,
    HEAT_DEG_DAYS = HTDD,
    COOL_DEG_DAYS = CLDD,
    HIGHEST_TEMP = EMXT,
    HIGHEST_DAY = EMXT_ATTRIBUTES,
    COOLEST_TEMP = EMNT,
    COOLEST_DAY = EMNT_ATTRIBUTES,
    GTE_32 = DX90,
    GTE_21 = DX70,
    GTE_00 = DX32,
    LTE_00 = DT32,
    LTE_N18 = DT00
  )

milanClimate$DATE <- lubridate::ym(milanClimate$DATE)

milanClimate$HIGHEST_DAY <- as.numeric(substr(gsub('\\D','',milanClimate$HIGHEST_DAY), nchar(gsub('\\D','',milanClimate$HIGHEST_DAY))-1, nchar(gsub('\\D','',milanClimate$HIGHEST_DAY))))

milanClimate$COOLEST_DAY <- as.numeric(substr(gsub('\\D','',milanClimate$COOLEST_DAY), nchar(gsub('\\D','',milanClimate$COOLEST_DAY))-1, nchar(gsub('\\D','',milanClimate$COOLEST_DAY))))

colnames(milanClimate)
```

## Marginal Summaries

Now that we have a cleaned data object, we can begin to plot charts to find and demonstrate different trends.

### Max Temperature and Minimum Temperature with Date

```{r}

subset_min <- milanClimate %>%
  filter(MEAN == min(MEAN, na.rm = TRUE)) %>%
  select(DATE, MEAN, MEAN_MIN, MEAN_MAX)
subset_min

subset_max <- milanClimate %>%
  filter(MEAN == max(MEAN, na.rm = TRUE)) %>%
  select(DATE, MEAN, MEAN_MIN, MEAN_MAX)
subset_max

subset_min_max <- bind_rows(subset_min, subset_max)
subset_min_max # Highest and Lowest Temperatures

ggplot(milanClimate, aes(x=DATE, y=MEAN)) +
  geom_bar(stat="identity", fill="blue", width=0.5) +
  geom_errorbar(aes(ymin = MEAN_MIN, ymax = MEAN_MAX), width = 0.2, position = position_dodge(width = 0.5)) +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Temperature Ranges by Date")

```

```{r}
# Monthly averages
ggplot(milanClimate, aes(x=DATE, y=MEAN)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Temperature by Date with Trend")
```

```{r}
# Extract year from the "date" column
new_data <- milanClimate %>%
  mutate(year = lubridate::year(DATE))

# Create a summary dataframe with yearly statistics
summary_df <- new_data %>%
  group_by(year) %>%
  summarise(
    MEAN = mean(MEAN, na.rm = TRUE),
    MIN = min(MEAN_MIN, na.rm = TRUE),
    MAX = max(MEAN_MAX, na.rm = TRUE)
  )
summary_df

# Print the summary dataframe
ggplot(summary_df, aes(x=year, y=MEAN)) +
  geom_bar(stat="identity", fill="blue", width=0.5) +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Annual Temperature Range with Trend")
```

```{r}

ggplot(milanClimate, aes(x=DATE, y=MEAN_MAX)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Maximum Temperature Annualy")

ggplot(milanClimate, aes(x=DATE, y=MEAN_MIN)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Temperature (C)") +
  ggtitle("Average Minimum Temperature Annualy")

ggplot(milanClimate, aes(x=DATE, y=HEAT_DEG_DAYS)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Heating Days")

ggplot(milanClimate, aes(x=DATE, y=COOL_DEG_DAYS)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Cooling Days")

ggplot(milanClimate, aes(x=DATE, y=GTE_32)) +
  geom_point(color='blue') +
  geom_smooth(method='lm', color='red') +
  labs(x="Date", y="Number of Days") +
  ggtitle("Temperatures about 32C")

```

## 2 Plans for Data Exploration

With the data, we plan to explore the shift of temperatures in the Milan area. Combining this with economic data from the area can share how the shift in temperature has aftected the economy. Or combining it with populaiton data, we can compare it with population shifts. Across the data, we should be able to find comparisons between the temperature and other local events.
